Cooperative Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-10 manuscript/paper.tex
paper: figures updates, general t2g case A,B matrices expansions calc and included in append.

\end{abstract}
\maketitle

@@

\az{The deformation $\delta$
will not couple these orbitals to each other either but 
modify the eg-t2g splitting as
${\Delta\to(1-\delta/3) \Delta }$,
and lift the degeneracies, as follows.
...
}
The crystal field Hamiltonian of the central B ion/JT ion
${H_{m',m''}^{JT} = (q_o/a)\sum_{l,m} V_{l,m}^{JT} C_{l}^{m'm''m} D_{l}}$
still turns out to be diagonal,
but the degeneracies of the two manifolds is lifted
and their average energies also shifted.
The $t_{2g}$ now splits into two levels,
with energies
\begin{align}
E_{xy} =&
\frac{32}{7} \left[
\left(1-\frac{1}{(\delta +1)^3}\right) D_2
-
\left(1-\frac{1}{(\delta +1)^5}\right) \frac{20}{9}D_4
\right]\frac{q_o}{a},\\
E_{yz} =& E_{zx} = -3/2 E_{xy},
\end{align}
as measured from their average 
%\begin{align}
$E_0^{t_{2g}} = -64q_oD_4 \left(2+(\delta +1)^{-5}\right)/9a$.
%\frac{64}{9} \left(-\frac{1}{(\delta +1)^5}-2\right) \text{D4}
%\end{align}
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&\frac{32}{7} \left[\left(1-\frac{1}{(\delta +1)^3}\right) \text{D2}+\frac{5}{3}\left(1-\frac{1}{(\delta +1)^5}\right) \text{D4}\right]\frac{q_o}{a},\\
E_{z^2}=&-E_{x^2-y^2},
\end{align}
measured from their average 
%\begin{align}
$E_0^{e_{g}} = 32 q_oD_4 \left(2+(\delta +1)^{-5}\right)/3a$.
%\frac{32}{3} \left(\frac{1}{(\delta +1)^5}+2\right) \text{D4}
%\end{align}
The crystal field splitting becomes
\begin{align}
\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=\frac{160}{9} \left(\frac{1}{(\delta +1)^5}+2\right) D_4.
\end{align}

At small $\delta$, 
we can expand the above expressions
 up to first order in $\delta$.
 $E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ are rescaled by ${1-5\delta/3}$,
 and we can write
\begin{align}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[(D_2-\frac{100}{27}  D_4)\right]\Delta_{\delta},\\
(E_{x^2-y^2},E_{z^2})&=(2,-2)
\left[(D_2+\frac{25}{9} D_4)\right]\Delta_{\delta}
\end{align}
where 
${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.

@@

\az{update the numbers,,, and quote realistic values of D2 and D4.}
Note the relative strengths of the two terms with $D_{2}$ and $D_{4}$.
\az{[refine argument]}
Typically, ${D_{2}\approx \sqrt{D_{4}}}$. 
Since
$D_{2}$ and $D_{4}$ are scaled with different 
powers of $a$ ($D_l=\braket{r^l}/a^l$),
their ratio depends on
the octahedron's size, i.e., $D_{2}/D_{4}\propto a^2$.
For well localised d-orbitals in a large octahedron,
${D_{2}/D_{4} > 25/9}$ so that elongation of the octahedron raises $xy$ and lowers $yz,xz$ states.
(Similarly, under such a condition, 
${D_{2}/D_{4} > 20/27}$ is automatically met so
it raises $x^2-y^2$ and lowers $z^2$).
However, if the d-orbitals are extended and octahedron is smaller,
a situation might result where ${20/27 < D_{2}/D_{4} < 25/9}$.
If so,
the ordering of the eg will be reversed as if a compression of the octahedron would have happened.
The extreme case of ${D_{2}/D_{4} < 20/27}$ (that seems less realistic)
would reverse both eg and t2g ordering for a given tetragonal distortion.

@@

@@

\com{
\begin{table}[htp]
\caption{default}
\begin{center}
\begin{tabular}{|L|L|}
\hline
V^A_{2,-2}& 64 \sqrt{\frac{3 \pi }{5}} \cos ^2\beta  \left(\frac{1}{\left(\cos ^2\alpha +2 \cos ^2\beta \right)^{5/2}}-\frac{\sec ^3\alpha }{(\cos 2 \beta +2)^{5/2}}\right)\\ \hline
V^A_{2,0}& 32 \sqrt{\frac{\pi }{5}} \left(\frac{2 \sec ^3\alpha  \sin ^2\beta }{(\cos 2 \beta +2)^{5/2}}-\frac{32 \sqrt{\frac{\pi }{5}} (\cos 2 \beta -\cos 2 \alpha )}{\left(\cos ^2\alpha +2 \cos ^2\beta \right)^{5/2}}\right)\\ \hline
V^A_{4,-2}& \frac{128}{3} \sqrt{5 \pi } \cos ^2\beta  \left(\frac{3 \cos 2 \alpha -\cos 2 \beta +2}{\left(\cos ^2\alpha +2 \cos ^2\beta \right)^{9/2}}+\frac{\sec ^5\alpha  (\cos 2 \beta -5)}{(\cos 2 \beta +2)^{9/2}}\right)\\ \hline
V^A_{4,0}& -\frac{16}{3} \sqrt{\pi } \left(\frac{8 \cos 2 \alpha  (3 \cos 2 \beta +2)-2 \cos 4 \alpha+12 \cos 2 \beta -3 \cos 4 \beta +9}{\left(\cos ^2\alpha +2 \cos ^2\beta \right)^{9/2}}\right. \\ 
&+\left. \frac{\sec ^5\alpha  (36 \cos 2 \beta -3 \cos 4 \beta +23)}{(\cos 2 \beta +2)^{9/2}}\right) \\ \hline
V^A_{4,4}&\frac{1}{3} (-128) \sqrt{35 \pi } \cos ^4\beta  \left(\frac{1}{\left(\cos ^2\alpha +2 \cos ^2\beta \right)^{9/2}}+\frac{\sec ^5\alpha }{(\cos 2 \beta +2)^{9/2}}\right)\\ \hline

\end{tabular}
\end{center}
\label{default}
\end{table}%
}

@@

\com{
\begin{table}[htp]
\caption{default}
\begin{center}
\begin{tabular}{|L|}
\hline
V^{B}_{2,-2}=  \frac{64 \sqrt{\frac{3 \pi }{5}} \sin ^2\alpha  \sec ^3\beta }{(\cos 2 \alpha +3)^{5/2}}\\ %\hline
V^{B}_{2,0}=  4 \sqrt{\frac{\pi }{5}} \left(\sec ^3\alpha -\frac{8 \sec ^3\beta }{(\cos 2 \alpha +3)^{3/2}}\right)\\ %\hline
 V^{B}_{4,-2}=  -\frac{128 \sqrt{5 \pi } \sin ^2\alpha  \sec ^5\beta }{3 (\cos 2 \alpha +3)^{7/2}}\\ %\hline
V^{B}_{4,0}= \frac{4}{3} \sqrt{\pi } \left(\frac{24 \sec ^5\beta }{(\cos 2 \alpha +3)^{5/2}}+\sec ^5\alpha \right)\\ %\hline
 V^{B}_{4,4}=  -\frac{16 \sqrt{35 \pi } (-20 \cos 2 \alpha +\cos 4 \alpha -13) \sec ^5\beta }{3 (\cos 2 \alpha +3)^{9/2}}\\ \hline
\end{tabular}
\end{center}
\label{default}
\end{table}%
}

@@

% from HA
\begin{align}
%\tilde H^A_{xy} =& -2 \Delta_{t2g},~
%\tilde H^A_{yz} = \tilde H^A_{zx} =\Delta_{t2g},\\
% \Delta_{t2g}(\alpha=0,\beta)
%\Delta_{t2g} =&\left( \Delta_{2}  D_2 + \Delta_4 D_4\right),\\
(E_{xy},&E_{yz},E_{zx})= (2, -1, -1)
\left[\frac{}{} \Delta_{2}  D_2 + \Delta_4 D_4\right],\\
\Delta_{l} =&  \left(q_A \Delta_{l}^A + q_B\Delta_{l}^B\right)/a,~(l=2,4),\\
\Delta_{2}^A =& 
%from A
\frac{64 \sin ^2\beta }{7 (\cos 2 \beta +2)^{5/2}},\\ % simplifies.... from:
 %\frac{16 (3 \sin \beta+\sin 3\beta)^2}{7 (\cos 2\beta+2)^{9/2}},\\
\Delta_4^A =&
 \frac{160 \sin ^2\beta}{63 (\cos 2\beta+2)^{9/2}} \times \\
&  (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82),\\
% \Delta_4^A =&
% \frac{160 \sin ^2\beta (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82)}{63 (\cos 2\beta+2)^{9/2}},\\
 %from B
\Delta_{2}^B =& \frac{2}{7} \left(1-\sec ^3\beta\right),\\
  \Delta_4^B =&\frac{5}{126} \left((7 \cos 4\beta-3) \sec ^5\beta-4\right),\\
 %\left( \Delta_{2}  D_2 + \Delta_4 D_4\right),\\
 \end{align}
along with an overall shift
\begin{align}
E_{0} =& \left(q_A E_{0}^A + q_BE_{0}^B\right)D_4/a,\\
E_{0}^A =&
 \frac{8 (92 \cos 2\beta+24 \cos 4\beta+20 \cos 6\beta+5 \cos 8\beta+51)}{9 (\cos 2\beta+2)^{9/2}},\\
E_{0}^B=&
 -\frac{1}{18} \left((5 \cos 4\beta+3) \sec ^5\beta+4\right),\\
 \end{align}

@@

$E_D$ and $E_{\pm}$ relative to their average ${E_0= \left(E_{xy} + 2 E_{yz}\right)/3}$
are shown in Fig.~\ref{fig:evals-alpha}.
$E_0$ itself shifts 
with the tilting $\alpha$, 
reflecting 
the change in eg-t2g gap $\Delta$.
This shift in $E_0$ is
%from its value 
%$\frac{D_4}{3} \left(\frac{512q}{81 \sqrt{3}}-2 (6-q)\right)$\\
%$\left(\frac{2}{729} \left(256 \sqrt{3}+243\right) q-4\right)D_4 $
%at $\alpha=0=\beta$, 
also included in Fig.~\ref{fig:evals-alpha}.

\com{
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/phi-0-q-0-r}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/phi-0-q-1-r}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/phi-0-q-2-r}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/phi-0-q-3-r}
\caption{d224=11.; norm should be eq to 0.00482334; d222=sqrt d224=0.0694503, d222/d224=14.3988 here. Dark state shuld b labelled, as well as pm states.
update to q=1,3 and analytic... d224norm=0.2?? or update theta0 figs to reduce d224 to 0.005?}
\label{default}
\end{center}
\end{figure}
}

@@

\com{
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/q-1-4-alpha-14deg-states-wts-vs-beta}
\caption{eigenstates: weights of xy, yz, zx.}
\label{default}
\end{center}
\end{figure}
}

@@

\begin{align}
E_0 &= \left(q_A E_0^A +q_B E_0^B\right)D_4/a,\\
E_0^A &=
\frac{256 \left(5 \left(4 \beta ^2+1\right) \alpha ^2+5 \beta ^2-3\right)}{243 \sqrt{3}}\\
E_0^B &=
(-\frac{1}{12} 5 \left(7 \beta ^2+4\right) \alpha ^2-\frac{5 \beta ^2}{3}+1).
\end{align}

@@

\com{
\begin{align}
\left(
\begin{array}{cc}
 -\frac{64 \left(\alpha ^2 \left(7 \beta ^2+3\right)-6 \beta ^2\right)}{189 \sqrt{3}} & \frac{640 \left(\alpha ^2 \left(\beta ^2+16\right)-32 \beta ^2\right)}{1701 \sqrt{3}} \\
 \frac{3}{14} \left(\alpha ^2 \left(3 \beta ^2+2\right)-4 \beta ^2\right) & \frac{1}{168} (-5) \left(\alpha ^2 \left(37 \beta ^2+46\right)-92 \beta ^2\right) \\
\end{array}
\right)
\end{align}

\begin{align}
mean=
\left(
\begin{array}{cc}
 0 & \frac{256 \left(5 \left(4 \beta ^2+1\right) \alpha ^2+5 \beta ^2-3\right)}{243 \sqrt{3}} \\
 0 & -\frac{1}{12} 5 \left(7 \beta ^2+4\right) \alpha ^2-\frac{5 \beta ^2}{3}+1 \\
\end{array}
\right)
\end{align}
}

@@

\begin{gather}
\nonumber
R=
\left(
\scalemath{0.9}{
\begin{array}{ccc}
 \frac{1}{2} ((\cos \alpha+1) \cos \beta+(\cos \alpha-1) \sin \beta) & \frac{1}{2} (\cos \beta-\sin \beta-\cos \alpha (\cos \beta+\sin \beta)) & \frac{\sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} \\
 \frac{1}{2} (-\cos \alpha \cos \beta+\cos \beta+\cos \alpha \sin \beta+\sin \beta) & \frac{1}{2} (\cos \beta+\cos \alpha (\cos \beta-\sin \beta)+\sin \beta) & \frac{\sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} \\
 -\frac{\sin \alpha}{\sqrt{2}} & \frac{\sin \alpha}{\sqrt{2}} & \cos \alpha \\
\end{array}
}
\right),\\ \nonumber
\com{
\mathcal{R}=
\left(
\scalemath{0.7}{
\begin{array}{ccccc}
 R_{12} R_{21}+R_{11} R_{22} & R_{13} R_{22}+R_{12} R_{23} & \sqrt{3} R_{13} R_{23} & R_{13} R_{21}+R_{11} R_{23} & R_{11} R_{21}-R_{12} R_{22} \\
 R_{22} R_{31}+R_{21} R_{32} & R_{23} R_{32}+R_{22} R_{33} & \sqrt{3} R_{23} R_{33} & R_{23} R_{31}+R_{21} R_{33} & R_{21} R_{31}-R_{22} R_{32} \\
 \frac{-1}{\sqrt{3}}\left(R_{11} R_{12}+R_{21} R_{22}-2 R_{31} R_{32}\right) & \frac{-1}{\sqrt{3}}\left(R_{12} R_{13}+R_{22} R_{23}-2 R_{32} R_{33}\right)  & -\frac{1}{2} R_{13}^2+R_{33}^2-\frac{R_{23}^2}{2} & \frac{-1}{\sqrt{3}}\left(R_{11} R_{13}+R_{21} R_{23}-2 R_{31} R_{33}\right) & 0\\ % identically 0 in our case of specific rotation and tilt.
 %\frac{1}{2\sqrt{3}}\left(-R_{11}^2+R_{12}^2-R_{21}^2+R_{22}^2+2 R_{31}^2-2 R_{32}^2\right) \\
 R_{12} R_{31}+R_{11} R_{32} & R_{13} R_{32}+R_{12} R_{33} & \sqrt{3} R_{13} R_{33} & R_{13} R_{31}+R_{11} R_{33} & R_{11} R_{31}-R_{12} R_{32} \\
 R_{11} R_{12}-R_{21} R_{22} & R_{12} R_{13}-R_{22} R_{23} & \frac{1}{2} \sqrt{3} \left(R_{13}^2-R_{23}^2\right) & R_{11} R_{13}-R_{21} R_{23} & \frac{1}{2} \left(R_{11}^2-R_{12}^2-R_{21}^2+R_{22}^2\right) \\
\end{array}
}
\right),\\
}
\mathcal{R}=
\left(
\scalemath{0.65}{
\begin{array}{ccccc}
 \frac{1}{4} (\cos 2 \alpha+3) \cos 2 \beta & \frac{\sin \alpha (\cos \alpha \cos 2 \beta+\sin 2 \beta)}{\sqrt{2}} & -\frac{1}{2} \sqrt{3} \cos 2 \beta \sin ^2\alpha & \frac{\sin \alpha (\sin 2 \beta-\cos \alpha \cos 2 \beta)}{\sqrt{2}} & \cos \alpha \sin 2 \beta \\
 \frac{\cos \alpha \sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} & \frac{1}{2} (\cos 2 \alpha (\cos \beta-\sin \beta)+\cos \alpha (\cos \beta+\sin \beta)) & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha (\sin \beta-\cos \beta) & \frac{1}{2} (\cos 2 \alpha (\sin \beta-\cos \beta)+\cos \alpha (\cos \beta+\sin \beta)) & -\frac{\sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} \\
 -\frac{1}{2} \sqrt{3} \sin ^2\alpha & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha & \frac{1}{4} (3 \cos 2 \alpha+1) & -\sqrt{\frac{3}{2}} \cos \alpha \sin \alpha & 0 \\
 \frac{\cos \alpha \sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} & \frac{1}{2} (\cos \alpha (\cos \beta-\sin \beta)-\cos 2 \alpha (\cos \beta+\sin \beta)) & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha (\cos \beta+\sin \beta) & \frac{1}{2} (\cos \alpha (\cos \beta-\sin \beta)+\cos 2 \alpha (\cos \beta+\sin \beta)) & \frac{\sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} \\
 -\left(\cos ^2\alpha+1\right) \cos \beta \sin \beta & \frac{\sin \alpha (\cos 2 \beta-2 \cos \alpha \cos \beta \sin \beta)}{\sqrt{2}} & \sqrt{3} \cos \beta \sin ^2\alpha \sin \beta & \frac{\sin \alpha (\cos 2 \beta+\cos \alpha \sin 2 \beta)}{\sqrt{2}} & \cos \alpha \cos 2 \beta \\
\end{array}
}
\right)
\end{gather}

\com{

\begin{align}
\mathcal{R}=&
\left(
\begin{array}{ccccc}
 R_{-1,1} R_{1,-1}+R_{-1,-1} R_{1,1} & R_{-1,0} R_{1,-1}+R_{-1,-1} R_{1,0} & \sqrt{3} R_{-1,0} R_{1,0} & R_{-1,1} R_{1,0}+R_{-1,0} R_{1,1} & R_{-1,1} R_{1,1}-R_{-1,-1} R_{1,-1} \\
 R_{-1,1} R_{0,-1}+R_{-1,-1} R_{0,1} & R_{-1,0} R_{0,-1}+R_{-1,-1} R_{0,0} & \sqrt{3} R_{-1,0} R_{0,0} & R_{-1,1} R_{0,0}+R_{-1,0} R_{0,1} & R_{-1,1} R_{0,1}-R_{-1,-1} R_{0,-1} \\
 -\frac{R_{-1,-1} R_{-1,1}-2 R_{0,-1} R_{0,1}+R_{1,-1} R_{1,1}}{\sqrt{3}} & -\frac{R_{-1,-1} R_{-1,0}-2 R_{0,-1} R_{0,0}+R_{1,-1} R_{1,0}}{\sqrt{3}} & -\frac{1}{2} R_{-1,0}^2+R_{0,0}^2-\frac{R_{1,0}^2}{2} & -\frac{R_{-1,0} R_{-1,1}-2 R_{0,0} R_{0,1}+R_{1,0} R_{1,1}}{\sqrt{3}} & \frac{R_{-1,-1}^2-R_{-1,1}^2-2 R_{0,-1}^2+2 R_{0,1}^2+R_{1,-1}^2-R_{1,1}^2}{2 \sqrt{3}} \\
 R_{0,1} R_{1,-1}+R_{0,-1} R_{1,1} & R_{0,0} R_{1,-1}+R_{0,-1} R_{1,0} & \sqrt{3} R_{0,0} R_{1,0} & R_{0,1} R_{1,0}+R_{0,0} R_{1,1} & R_{0,1} R_{1,1}-R_{0,-1} R_{1,-1} \\
 R_{1,-1} R_{1,1}-R_{-1,-1} R_{-1,1} & R_{1,-1} R_{1,0}-R_{-1,-1} R_{-1,0} & \frac{1}{2} \sqrt{3} \left(R_{1,0}^2-R_{-1,0}^2\right) & R_{1,0} R_{1,1}-R_{-1,0} R_{-1,1} & \frac{1}{2} \left(R_{-1,-1}^2-R_{-1,1}^2-R_{1,-1}^2+R_{1,1}^2\right) \\
\end{array}
\right)
\end{align}

>>>>

\end{abstract}
%\maketitle

@@

\az{The deformation $\delta$
will not couple these orbitals to each other either but 
modify the eg-t2g splitting as
${\Delta\to(1-\delta/3) \Delta }$,
and lift the degeneracies, as follows.
...
}
The crystal field Hamiltonian of the central B ion/JT ion
${H_{m',m''}^{JT} = (q_o/a)\sum_{l,m} V_{l,m}^{JT} C_{l}^{m'm''m} D_{l}}$
still turns out to be diagonal,
but the degeneracies of the two manifolds is lifted
and their average energies also shifted.
Defining 
$F_n(\delta) = 1-1/(1+\delta)^n$,
we see that
the $t_{2g}$ now splits into two levels,
with energies

\begin{align}
E_{xy} =&
\frac{32}{7}\frac{q_o}{a} \left(
 D_2F_3 - \frac{20}{9}D_4F_5 
\right),\\
E_{yz} =& E_{zx} = -\frac{1}{2} E_{xy},
\end{align}
as measured from their average 
%\begin{align}
%$E_0^{t_{2g}} = -64q_oD_4 \left(2+(\delta +1)^{-5}\right)/9a$.
$E_0^{t_{2g}} = -64q_oD_4 \left(3-F_5\right)/9a$.
%\frac{64}{9} \left(-\frac{1}{(\delta +1)^5}-2\right) \text{D4}
%\end{align}
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&
\frac{32}{7}\frac{q_o}{a} \left(D_2F_3+\frac{5}{3} D_4F_5\right),\\
E_{z^2}=&-E_{x^2-y^2},\\
\end{align}
measured from their average 
%\begin{align}
%$E_0^{e_{g}} = 32 q_oD_4 \left(2+(\delta +1)^{-5}\right)/3a$.
$E_0^{e_{g}} = 32 q_oD_4 \left(3-F_5\right)/3a$.
%\frac{32}{3} \left(\frac{1}{(\delta +1)^5}+2\right) \text{D4}
%\end{align}
The crystal field splitting becomes
%\begin{align}
$\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=
%\frac{160}{9} \left(\frac{1}{(\delta +1)^5}+2\right) D_4.
160q_oD_4\left(3-F_5\right)/9a$.
%\end{align}
At small $\delta$, 
expanding 
$F_3,F_5$
 up to first order in $\delta$,
we obtain
\begin{align}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[D_2-\frac{100}{27}  D_4\right]\frac{48}{7}\frac{q_o}{a}\delta,\\
(E_{x^2-y^2},E_{z^2})&=(1,-1)
\left[D_2+\frac{25}{9} D_4\right]\frac{96}{7}\frac{q_o}{a}\delta,
\end{align}
whereas
%${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.
$E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ are rescaled by a factor of ${1-5\delta/3}$.

@@

\az{update the numbers,,, and quote realistic values of D2 and D4.:
$D_2/D_4 \simeq 1\sqrt{D_4} \simeq 15-25$ estimated from realistic field splittings, 
$q_o=2,a=7$ bohr,
$\Delta=0.5-3$ eV. ... check... 
}
Note the relative strengths of the two terms with $D_{2}$ and $D_{4}$.
\az{[refine argument]}
Typically, ${D_{2}\approx \sqrt{D_{4}}}$. 
Since
$D_{2}$ and $D_{4}$ are scaled with different 
powers of $a$ ($D_l=\braket{r^l}/a^l$),
their ratio depends on
the octahedron's size, i.e., $D_{2}/D_{4}\propto a^2$.
For well localised d-orbitals in a large octahedron,
${D_{2}/D_{4} > 25/9}$ so that elongation of the octahedron raises $xy$ and lowers $yz,xz$ states.
(Similarly, under such a condition, 
${D_{2}/D_{4} > 20/27}$ is automatically met so
it raises $x^2-y^2$ and lowers $z^2$).
However, if the d-orbitals are extended and octahedron is smaller,
a situation might result where ${20/27 < D_{2}/D_{4} < 25/9}$.
If so,
the ordering of the eg will be reversed as if a compression of the octahedron would have happened.
The extreme case of ${D_{2}/D_{4} < 20/27}$ (that seems less realistic)
would reverse both eg and t2g ordering for a given tetragonal distortion.

@@

\com{
\begin{align}
E_{xy} =&
\frac{32}{7} \left[
\left(1-\frac{1}{(\delta +1)^3}\right) D_2
-
\left(1-\frac{1}{(\delta +1)^5}\right) \frac{20}{9}D_4
\right]\frac{q_o}{a},\\
E_{yz} =& E_{zx} = -3/2 E_{xy},
\end{align}
as measured from their average 
%\begin{align}
$E_0^{t_{2g}} = -64q_oD_4 \left(2+(\delta +1)^{-5}\right)/9a$.
%\frac{64}{9} \left(-\frac{1}{(\delta +1)^5}-2\right) \text{D4}
%\end{align}
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&\frac{32}{7} \left[\left(1-\frac{1}{(\delta +1)^3}\right) \text{D2}+\frac{5}{3}\left(1-\frac{1}{(\delta +1)^5}\right) \text{D4}\right]\frac{q_o}{a},\\
E_{z^2}=&-E_{x^2-y^2},
\end{align}
measured from their average 
%\begin{align}
$E_0^{e_{g}} = 32 q_oD_4 \left(2+(\delta +1)^{-5}\right)/3a$.
%\frac{32}{3} \left(\frac{1}{(\delta +1)^5}+2\right) \text{D4}
%\end{align}
The crystal field splitting becomes
\begin{align}
\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=\frac{160}{9} \left(\frac{1}{(\delta +1)^5}+2\right) D_4.
\end{align}
}

@@

@@

@@

% leading order expansion in beta:
%A 2,4: \left\{\frac{64 \beta ^2}{63 \sqrt{3}}+O\left(\beta ^3\right),\frac{40960 \beta ^2}{5103 \sqrt{3}}+O\left(\beta ^3\right)\right\}
%B 2,4: \left\{-\frac{3 \beta ^2}{7}+O\left(\beta ^3\right),-\frac{115 \beta ^2}{63}+O\left(\beta ^3\right)\right\}

% from HA
\begin{align}
%\tilde H^A_{xy} =& -2 \Delta_{t2g},~
%\tilde H^A_{yz} = \tilde H^A_{zx} =\Delta_{t2g},\\
% \Delta_{t2g}(\alpha=0,\beta)
%\Delta_{t2g} =&\left( \Delta_{2}  D_2 + \Delta_4 D_4\right),\\
(E_{xy},&E_{yz},E_{zx})= (2, -1, -1)
\left[\frac{}{} \Delta_{2}  D_2 + \Delta_4 D_4\right],\\
\Delta_{l} =&  \left(q_A \Delta_{l}^A + q_B\Delta_{l}^B\right)/a,~(l=2,4),\\
\Delta_{2}^A =& 
%from A
\frac{64 \sin ^2\beta }{7 (\cos 2 \beta +2)^{5/2}} % simplifies.... from:
 %\frac{16 (3 \sin \beta+\sin 3\beta)^2}{7 (\cos 2\beta+2)^{9/2}},\\
 \simeq \frac{64}{63 \sqrt{3}}\beta ^2,\\
\Delta_4^A =&
 \frac{160 \sin ^2\beta}{63 (\cos 2\beta+2)^{9/2}} \times \\
&  (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82),\\
\simeq& \frac{40960}{5103 \sqrt{3}}\beta ^2,\\
% \Delta_4^A =&
% \frac{160 \sin ^2\beta (125 \cos 2\beta+42 \cos 4\beta+7 \cos 6\beta+82)}{63 (\cos 2\beta+2)^{9/2}},\\
 %from B
\Delta_{2}^B =& \frac{2}{7} \left(1-\sec ^3\beta\right)
\simeq -\frac{3}{7}\beta ^2,\\
  \Delta_4^B =&\frac{5}{126} \left((7 \cos 4\beta-3) \sec ^5\beta-4\right)
  \simeq -\frac{115}{63}\beta ^2,\\
 %\left( \Delta_{2}  D_2 + \Delta_4 D_4\right),\\
 \end{align}
along with an overall shift
\begin{align}
E_{0}^{t_{2g}} =& \left(q_A E_{0}^A + q_BE_{0}^B\right)D_4/a,\\
E_{0}^A =&
 \frac{8 (92 \cos 2\beta+24 \cos 4\beta+20 \cos 6\beta+5 \cos 8\beta+51)}{9 (\cos 2\beta+2)^{9/2}},\\
 \simeq& \frac{512}{243 \sqrt{3}}-\frac{2560 \beta ^2}{729 \sqrt{3}},\\
E_{0}^B=&
 -\frac{1}{18} \left((5 \cos 4\beta+3) \sec ^5\beta+4\right)
 \simeq -\frac{2}{3}+\frac{10 \beta ^2}{9}.
 \end{align}

@@

$E_D$ and $E_{\pm}$ relative to their average ${E_0^{t_{2g}}= \left(E_{xy} + 2 E_{yz}\right)/3}$
are shown in Fig.~\ref{fig:evals-alpha}.
$E_0^{t_{2g}}$ itself shifts 
with the tilting $\alpha$, 
reflecting 
the change in eg-t2g gap $\Delta$.
This shift in $E_0^{t_{2g}}$ is
%from its value 
%$\frac{D_4}{3} \left(\frac{512q}{81 \sqrt{3}}-2 (6-q)\right)$\\
%$\left(\frac{2}{729} \left(256 \sqrt{3}+243\right) q-4\right)D_4 $
%at $\alpha=0=\beta$, 
also included in Fig.~\ref{fig:evals-alpha}.

@@

@@

\begin{align}
E_0^{e_{g}} &= \left(q_A E_0^A +q_B E_0^B\right)D_4/a,\\
E_0^A &=
\frac{256 \left(5 \left(4 \beta ^2+1\right) \alpha ^2+5 \beta ^2-3\right)}{243 \sqrt{3}}\\
E_0^B &=
(-\frac{1}{12} 5 \left(7 \beta ^2+4\right) \alpha ^2-\frac{5 \beta ^2}{3}+1).
\end{align}

@@

@@

\begin{gather}
\nonumber
R=
\left(
\scalemath{0.9}{
\begin{array}{ccc}
 \frac{1}{2} ((\cos \alpha+1) \cos \beta+(\cos \alpha-1) \sin \beta) & \frac{1}{2} (\cos \beta-\sin \beta-\cos \alpha (\cos \beta+\sin \beta)) & \frac{\sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} \\
 \frac{1}{2} (-\cos \alpha \cos \beta+\cos \beta+\cos \alpha \sin \beta+\sin \beta) & \frac{1}{2} (\cos \beta+\cos \alpha (\cos \beta-\sin \beta)+\sin \beta) & \frac{\sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} \\
 -\frac{\sin \alpha}{\sqrt{2}} & \frac{\sin \alpha}{\sqrt{2}} & \cos \alpha \\
\end{array}
}
\right),\\ \nonumber
\mathcal{R}=
\left(
\scalemath{0.65}{
\begin{array}{ccccc}
 \frac{1}{4} (\cos 2 \alpha+3) \cos 2 \beta & \frac{\sin \alpha (\cos \alpha \cos 2 \beta+\sin 2 \beta)}{\sqrt{2}} & -\frac{1}{2} \sqrt{3} \cos 2 \beta \sin ^2\alpha & \frac{\sin \alpha (\sin 2 \beta-\cos \alpha \cos 2 \beta)}{\sqrt{2}} & \cos \alpha \sin 2 \beta \\
 \frac{\cos \alpha \sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} & \frac{1}{2} (\cos 2 \alpha (\cos \beta-\sin \beta)+\cos \alpha (\cos \beta+\sin \beta)) & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha (\sin \beta-\cos \beta) & \frac{1}{2} (\cos 2 \alpha (\sin \beta-\cos \beta)+\cos \alpha (\cos \beta+\sin \beta)) & -\frac{\sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} \\
 -\frac{1}{2} \sqrt{3} \sin ^2\alpha & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha & \frac{1}{4} (3 \cos 2 \alpha+1) & -\sqrt{\frac{3}{2}} \cos \alpha \sin \alpha & 0 \\
 \frac{\cos \alpha \sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} & \frac{1}{2} (\cos \alpha (\cos \beta-\sin \beta)-\cos 2 \alpha (\cos \beta+\sin \beta)) & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha (\cos \beta+\sin \beta) & \frac{1}{2} (\cos \alpha (\cos \beta-\sin \beta)+\cos 2 \alpha (\cos \beta+\sin \beta)) & \frac{\sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} \\
 -\left(\cos ^2\alpha+1\right) \cos \beta \sin \beta & \frac{\sin \alpha (\cos 2 \beta-2 \cos \alpha \cos \beta \sin \beta)}{\sqrt{2}} & \sqrt{3} \cos \beta \sin ^2\alpha \sin \beta & \frac{\sin \alpha (\cos 2 \beta+\cos \alpha \sin 2 \beta)}{\sqrt{2}} & \cos \alpha \cos 2 \beta \\
\end{array}
}
\right)
\end{gather}
